\bf Supervised Tensor Decomposition with Interactive Side Information

### 2020-09-22 tensor_regress/JCGS_submission/jcgs-tensor/jcgs-tensor.tex
jcgs
update

\newcommand{\blind}{1}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

\theoremstyle{definition}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}{Lemma}
\newtheorem{prop}{Proposition}
\newtheorem{pro}{Property}
\newtheorem{cor}{Corollary}[section]

@@

@@

\if1\blind
{
  \title{\bf Supervised Tensor Decomposition with Interactive Side Information}
  % blind version
  \author{Jiaxin Hu, Zhuoyan Xu, Chanwoo Lee, and Miaoyan Wang\\
    Department of Statistics, University of Wisconsin-Madison}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Supervised Tensor Decomposition with Interactive Side Information}
\end{center}
  \medskip
} \fi

@@

\noindent%
{\it Keywords:} Applications and case studies, Tensor data analysis, Supervised dimension reduction, Exponential family distribution, Generalized multilinear model
\vfill

@@

In addition to the aforementioned challenges, many tensor datasets consist of non-Gaussian measurements. Examples include the political interaction dataset \citep{nickel2011three} which measures action counts between countries under various relations, and the brain connectivity network dataset \citep{zhang2018mapping} which is a collection of binary adjacency matrices. Classical tensor decomposition methods are based on minimizing the Frobenius norm of deviation, leading to suboptimal predictions for binary- or count-valued response variables. A number of supervised tensor methods have been proposed \citep{narita2012tensor, zhao2012higher, yu2016learning,lock2018supervised} to address the tensor regression problem in various forms, such as scalar-to-tensor regression and tensor-response regression. These methods often assume Gaussian distribution for the tensor entries, or impose random designs for the feature matrices, both of which are less suitable for applications of our interest. The gap between theory and practice means a great opportunity to modeling paradigms and better capture the complexity in tensor data.

\begin{figure*}[t]
\begin{center}
\includegraphics[width=16cm]{demo.pdf}
\vspace{-.5cm}
\end{center}
\caption{Examples of supervised tensor decomposition with interactive side information. (a) Network population model. (b) Spatio-temporal growth model.} \label{fig:intro1}
\vspace{-.3cm}
\end{figure*}

@@

Our work is closely related to but also clearly distinctive from several lines of previous work. The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor, hong2020generalized} that aims to find the best low-rank representation of a data tensor. In contrast, our model is a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and multiple features. The low-rank factorization is determined jointly by the tensor data and feature matrices in our model.

The second line of work studies the tensor-on-tensor regression~\citep{raskutti2019convex, lock2018tensor,gahrooei2020multiple}. Our model shares a common ground with earlier approaches, but we provide more efficient solutions to new settings that have more practical significance. As we show in Section~\ref{sec:connection}, the supervised tensor decomposition has an interesting connection with tensor-on-tensor regression. Previous methods~\citep{lock2018tensor} mainly focus on Gaussian tensors. The Frobenius norm used in the objective function is statistical suboptimal for general exponential family tensors. Maximum likelihood estimator (MLE) is studied in~\citep{raskutti2019convex} and a convex relaxation algorithm is proposed to solve for low-rank tensor coefficients. However, in the tensor case, convex MLE suffers from both computational intractability and statistical suboptimality. We advocate a non-convex approach and provide strong evidence for its success in our setting. Most previous tensor regression focuses on prediction~\citep{lock2018tensor,raskutti2019convex}, and we go step further by finding the sufficient dimension reduction~\citep{adragni2009sufficient}, $\textup{Span}(\mM_k)$, that facilitates the identification of interaction effects in features (see Figure~\ref{fig:intro1}b). The latter approach greatly improves the \emph{interpretability} in prediction. In this regards, our method opens up new opportunities for tensor data analysis in a wider range of applications.

The third line of work uses side information for various tensor learning tasks, such as for completion~\citep{narita2012tensor} and for recommendation system~\citep{ioannidis2019coupled,farias2019learning}. These methods also study tensors with side information, but they take regularization approaches to penalize predictions that are distant from side information~\cite{lock2018supervised}. One important difference is that their goal is prediction but not parameter estimation. The effects of features and their interactions are not estimated in these data-driven approaches. In contrast, our goal is interpretable prediction, and we estimate factor matrices $\mM_k$ using a model-based approach. Estimating $\mM_k$ allows us to identify sufficient features and the interactions thereof. We numerically compare these two approaches in Section~\ref{sec:simulation}.

The paper is organized as follows. Section~\ref{sec:pre} introduces tensor preliminaries. Section~\ref{sec:model} presents the main model and three motivating examples for supervised tensor decomposition. 
We describe the quasi-likelihood estimation in Section~\ref{sec:est} and provide numerical experiments in Section~\ref{sec:simulation}. In Section~\ref{sec:data}, we present two data applications, one on diffusion tensor imaging data and the other on multi-relational network data.
%and applications to diffusion tensor imaging data and multi-relational social network data in Section~\ref{sec:data}. 
%In Section~\ref{sec:simulation}, we present numerical experiments and assess the performance in comparison to alternative methods. In Section~\ref{sec:data}, we apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We conclude in Section~\ref{sec:con} with discussions about our findings and avenues of future work.

\section{Preliminaries}\label{sec:pre}
We introduce the basic tensor properties~\citep{kolda2009tensor} used in the paper. We use lower-case letters (e.g.,\ $a,b,c$) for scalars and vectors, upper-case boldface letters (e.g.,\ $\mA, \mB, \mC$) for matrices, and calligraphy letters (e.g.,\ $\tA, \tB, \tC$) for tensors of order three or greater. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as $\tY \times_1\mX_1\times \cdots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\cdots x_{j_K,i_K}^{(K)}}$,
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use the shorthand $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as $\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}.$
The tensor Frobenius norm is defined as $\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}$, and the maximum norm is defined $\mnormSize{}{\tY}=\max_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}.$
When $a$ is a vector, we use $\vnormSize{}{a}=\langle a,a \rangle^{1/2}$ to denote the vector 2-norm. A higher-order tensor can be reshaped into a lower-order object. We use $\textup{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ to denote the operation that reshapes the tensor along mode $k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The multilinear rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\textup{Unfold}_k(\tY)$, $k=1,\ldots,K$. We use $\sigma_{\min}(\cdot)$ and $\sigma_{\max}(\cdot)$ denote the smallest and largest matrix singular values. We let $\mI_d$ denote the $d \times d$ identity matrix and $[d]$ denote the $d$-set $\{1,\ldots,d\}$. For ease of notation, we allow the basic arithmetic operators (e.g.,\ $+, -, \geq $) and univariate functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner. 
%We use $f'(\cdot)$, $f''(\cdot)$ to denote the first-order and second-order derivative of a function $f(\cdot)$.

@@

\subsection{Three examples}
We give three seemingly different examples that can all be formulated as our supervised tensor decomposition model~\eqref{eq:decomp}.
\begin{example}[Spatio-temporal growth model]
The growth curve model~\citep{gabriel1998generalised} was originally proposed as an example of bilinear model for matrix data, and we adopt its higher-order extension here. Let $\tY=\entry{y_{ijk}}\in\mathbb{R}^{d \times m\times n}$ denote the pH measurements of $d$ lakes at $m$ levels of depth and for $n$ time points. Suppose the sampled lakes belong to $q$ types, with $p$ lakes in each type. Let $\{\ell_j\}_{j\in[m]}$ denote the sampled depth levels and $\{t_k\}_{k\in[n]}$ the time points. Assume that the expected pH trend in depth is a polynomial of order at most $r$ and that the expected trend in time is a polynomial of order $s$. Then, the conditional mean model for the spatio-temporal growth can be represented as
\begin{equation}\label{eq:time}
\mathbb{E}(\tY|\mX_1,\mX_2,\mX_3)=\tC\times\{\mX_1\mM_1,\ \mX_2\mM_2,\ \mX_3\mM_3\},
\end{equation}
where $\mX_1=\text{blockdiag}\{\mathbf{1}_p,\ldots,\mathbf{1}_p\}\in \{0,1\}^{d\times q}$ is the design matrix for lake types, and
\[
\mX_2=
\begin{pmatrix}
1 & \ell_1&\cdots &\ell^{r}_1\\
1 & \ell_2&\cdots &\ell^{r}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&\ell_{m}&\cdots&\ell^{r}_{m}
\end{pmatrix},\quad
\mX_3=
\begin{pmatrix}
1 & t_1&\cdots &t^{s}_1\\
1 & t_2&\cdots &t^{s}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&t_{n}&\cdots&t^{s}_{n}
\end{pmatrix}
\]
are the design matrices for spatial and temporal effects, respectively, $\tC\in\mathbb{R}^{r_1\times r_2\times r_3}$ is the unknown core tensor, and $\mM_k$ are unknown dimension reduction matrices on each mode. The factors $\mX_k\mM_k$ are sufficient features in the mean model~\eqref{eq:time}. The spatial-temporal model is a special case of our supervised tensor decomposition model~\eqref{eq:decomp}, with features available on each of the three modes.
\end{example}

@@

\subsection{Connection to sufficient dimension reduction and tensor-on-tensor regression}\label{sec:connection}
One important implication of our method is that we allow high-dimensionality in both tensor dimension $d_k$ and feature dimension $p_k$. Recall that the model rank $r_k$ is typically smaller than $d_k$ and $p_k$. In such a case, the matrices $\mM_k$ serve the role of simultaneous dimension reduction of the data tensor and features. The role of the sufficient features $\mX_k\mM_k$ can be seen in the following two conditional independence assumptions in model~\eqref{eq:decomp},

\begin{align}
\tY \ci \{\mX_k\}&\ \big|\ \{\mX_k\mM_k\} \ \ \text{(independence between the tensor and features)},\\
y_{i_1,\ldots,i_K} \ci y_{i'_1,\ldots,i'_K} & \ \big|\ \{\mX_k\mM_k\}  \ \ \text{(independence within the tensor)},
\end{align}
where the second line holds for all $(i_1,\ldots,i_K)\neq(i'_1,\ldots,i'_K) \in[d_1]\times\cdots\times[d_K]$, and $\ci$ denotes the independence. The first property highlights the ``decorrelation'' role of $\mM_k$, in the same spirit as the sufficient dimension reduction~\citep{adragni2009sufficient} in supervised learning, whereas the second property highlights the tensor dimension reduction in the usual unsupervised sense (consider, for example, $\mX_k=\mathbf{I}_{d_k}$ for all $k\in[K]$).

Our model also has a close connection with tensor-on-tensor regression~\citep{raskutti2019convex, lock2018tensor,gahrooei2020multiple}. Specifically, model~\eqref{eq:decomp} can be viewed as a multivariate regression model, where the response is the vectorized tensor and the covariates are interactions between sufficient features across modes. We take an order-3 tensor under the Gaussian model for illustration. Let $\mX,\mZ,\mW$ denote the feature matrix on mode $k=1, 2, 3$, respectively. Suppose that each mode has two-dimensional sufficient features, denoted $\mM_1\mX=[\mx_1,\mx_2]$, $\mM_2\mZ=[\mz_1,\mz_2]$, $\mM_3\mW=[\mw_1,\mw_2]$. Here $\mx_1,\mx_2,\ldots,\mw_1, \mw_2$ are column vectors. Then the model~\eqref{eq:decomp} is a regression model with across-mode interactions,
\begin{equation}\label{eq:interaction}
\mathbb{E}(y_{ijk}|\mX,\mZ,\mW)=c_{111}\mx_{1i}\mz_{1j}\mw_{1k}+c_{121}\mx_{i1}\mz_{j2}\mw_{k1}+\cdots+c_{221}\mx_{2i}\mz_{2j}\mw_{1k}+c_{222}\mx_{2i}\mz_{2j}\mw_{2k},
\end{equation}
where $\entry{c_{ijk}} \in \mathbb{R}^{2\times 2\times 3}$ are unknown interaction effects, $\mx_{1i}$ denotes the $i$-th entry in the feature vector $\mx_1$, and similar notations apply to other features. Note that lower-order interactions are naturally incorporated in~\eqref{eq:interaction} if we include an intercept column in the sufficient feature matrices. Model~\eqref{eq:interaction} shows the benefit of our supervised tensor decomposition for identifying across-mode interactions.

@@

\begin{table}[htb]
\centering
\begin{tabular}{c|ccc}
Data type &Gaussian & Poisson& Bernoulli\\
\hline
Domain $\mathbb{Y}$& $\mathbb{R}$&$\mathbb{N}$&$\{0,1\}$\\
 $b(\theta)$&$\theta^2/2$& $\exp(\theta)$&$\log (1+\exp(\theta))$\\
 link $f(\theta)$&$\theta$&$\exp(\theta)$&$(1+\exp(-\theta))^{-1}$
\end{tabular}
\vspace{-.2cm}
\caption{Canonical links for common distributions.}\label{table:link}
\end{table}

In our context, we model the tensor entries $y_{i_1,\ldots,i_K}$, conditional on $\theta_{i_1,\ldots,i_K}$, as independent draws from an exponential family. Ignoring constants that do not depend on $\Theta$, the quasi log-likelihood of~\eqref{eq:decomp} is equal to Bregman distance between $\tY$ and $b'(\Theta)$:
\begin{equation}\label{eq:loglikelihood}
\small
\tL_{\tY}(\tC,\mM_1,\ldots,\mM_K)=\langle \tY, \Theta \rangle - \sum\nolimits_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K})\ \text{with}\ \Theta=\tC\times\{\mM_1\mX_1,\ldots,\mM_K\mX_K\}.
\end{equation}
We propose a constrained maximum quasi-likelihood estimator (M-estimator),
\begin{align} \label{eq:MLE} 
(\hat \tC, \hat \mM_1,\ldots,\hat \mM_k) &=\argmax_{(\tC,\mM_1,\ldots,\mM_K)\in \tP} \ \tL_{\tY}(\tC,\mM_1,\ldots,\mM_K),
\end{align}
where the parameter space $\tP$ is
\[
\tP=\left\{(\tC, \mM_1,\ldots,\mM_K) \ \Big| \ \mM_k^T\mM_k=\mathbf{I}_{r_k},\ \mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha \right\}.
\]
The maximum norm constraint on the linear predictor $\Theta$ is a technical condition to avoid the divergence in the non-Gaussian variance.

@@

We conclude this section by providing the accuracy measured in KL divergence.
%, for the response distribution.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=16cm]{compare_GLM.pdf}
\caption{Performance comparison under stochastic block models. The three panels plot the MSE when the response tensors are generated from (a) Gaussian (b) Poisson and (c) Bernoulli models. The $x$-axis represents the number of blocks in the networks. }\label{fig:glm}
\vspace{-.2cm}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=15cm]{compare_alternative.pdf} 
\caption{Comparison between different tensor methods. Panels (a) and (b) plot MSPE versus the number of modes with available features. Panels (c) and (d) plot MSPE versus the effective sample size $d^2$.}~\label{fig:compare}
%We consider rank $\mr=(3,3,3)$ (low), $\mr=(4,5,6)$ (high), and signal $\alpha =3 $ (low), $\alpha=6$ (high).}~\label{fig:compare}
\vspace{-.4cm}
\end{figure}

>>>>

%\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\theoremstyle{definition}
\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}{Lemma}
\newtheorem{prop}{Proposition}[section]
\newtheorem{pro}{Property}
\newtheorem{cor}{Corollary}[section]

@@

%\bibliographystyle{natbib}

@@

\if0\blind
{
  \title{\bf Supervised Tensor Decomposition with Interactive Side Information}
  \author{Jiaxin Hu, Zhuoyan Xu, Chanwoo Lee, and Miaoyan Wang\thanks{The authors gratefully acknowledge NSF grant DMS-1915978 and funding from the Wisconsin Alumni Research Foundation. }\hspace{.2cm}\\
    Department of Statistics, University of Wisconsin-Madison}
  \maketitle

} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Supervised Tensor Decomposition with Interactive Side Information}
\end{center}
  \medskip
} \fi

@@

\noindent%
{\it Keywords:}  Tensor data analysis, Supervised dimension reduction, Exponential family distribution, Generalized multilinear model, Alternating optimization
\vfill

@@

In addition to the aforementioned challenges, many tensor datasets consist of non-Gaussian measurements. Examples include the political interaction dataset \citep{nickel2011three,hu2015scalable} which measures action counts between countries under various relations, and the brain connectivity network dataset \citep{zhang2018mapping,wang2018learning,wang2019common} which is a collection of binary adjacency matrices. Classical tensor decomposition methods are based on minimizing the Frobenius norm of deviation, leading to suboptimal predictions for binary- or count-valued response variables. A number of supervised tensor methods have been proposed \citep{narita2012tensor, lock2018supervised,rai2014scalable} to address the tensor regression problem in various forms, such as scalar-to-tensor regression and tensor-response regression. These methods often assume Gaussian distribution for the tensor entries, or impose random designs for the feature matrices, both of which are less suitable for applications of our interest. The gap between theory and practice means a great opportunity to modeling paradigms and better capture the complexity in tensor data.

\begin{figure*}[t]
\begin{center}
\includegraphics[width=16cm]{demo.pdf}
\end{center}
\caption{Examples of supervised tensor decomposition with interactive side information. (a) Network population model. (b) Spatio-temporal growth model.} \label{fig:intro1}
\end{figure*}

@@

Our work is closely related to but also clearly distinctive from several lines of previous work. The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor, hong2020generalized,wang2017tensor,bi2018recom,chi2012tensors} that aims to find the best low-rank representation of a data tensor. In contrast, our model is a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and multiple features. The low-rank factorization is determined jointly by the tensor data and feature matrices in our model.

The second line of work studies the tensor-on-tensor regression~\citep{raskutti2019convex, lock2018tensor,gahrooei2020multiple}. Our model shares a common ground with earlier approaches, but we provide more efficient solutions to new settings that have more practical significance. As we show in Section~\ref{sec:connection}, the supervised tensor decomposition has an interesting connection with tensor-on-tensor regression. Previous methods~\citep{lock2018tensor,lock2018supervised} mainly focus on Gaussian tensors. The Frobenius norm used in the objective function is statistical suboptimal for general exponential family tensors. Maximum likelihood estimator (MLE) is studied in~\cite{raskutti2019convex} and a convex relaxation algorithm is proposed to solve for low-rank tensor coefficients. However, in the tensor case, convex MLE suffers from both computational intractability and statistical suboptimality. We advocate a non-convex approach and provide strong evidence for its success in our setting. Most previous tensor regression focuses on prediction~\citep{lock2018tensor,raskutti2019convex}, and we go step further by finding the sufficient dimension reduction~\citep{adragni2009sufficient}, $\textup{Span}(\mM_k)$, that facilitates the identification of interaction effects in features (see Figure~\ref{fig:intro1}b). The latter approach greatly improves the \emph{interpretability} in prediction. In this regards, our method opens up new opportunities for tensor data analysis in a wider range of applications.

The third line of work uses side information for various tensor learning tasks, such as for completion~\citep{narita2012tensor,song2019tensor,cao2016semi} and for recommendation system~\citep{ioannidis2019coupled,farias2019learning}. These methods also study tensors with side information, but they take regularization approaches to penalize predictions that are distant from side information~\citep{cao2016semi,song2019tensor}. One important difference is that their goal is prediction but not parameter estimation. The effects of features and their interactions are not estimated in these data-driven approaches. In contrast, our goal is interpretable prediction, and we estimate factor matrices $\mM_k$ using a model-based approach. Estimating $\mM_k$ allows us to identify sufficient features and the interactions thereof. We numerically compare these two approaches in Section~\ref{sec:simulation}.

The remainder of the paper is organized as follows. Section~\ref{sec:pre} introduces tensor preliminaries. Section~\ref{sec:model} presents the main model and three motivating examples for supervised tensor decomposition. We describe the quasi-likelihood estimation and alternating optimization algorithm in Section~\ref{sec:est}. In Section~\ref{sec:simulation}, we present numerical experiments and assess the performance in comparison to alternative methods. In Section~\ref{sec:data}, we apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We conclude in Section~\ref{sec:con} with discussions about our findings and avenues of future work. All proofs are deferred to Supplementary Materials.

\section{Preliminaries}\label{sec:pre}
We introduce the basic tensor properties~\citep{kolda2009tensor} used in the paper. We use lower-case letters (e.g.,\ $a,b,c$) for scalars and vectors, upper-case boldface letters (e.g.,\ $\mA, \mB, \mC$) for matrices, and calligraphy letters (e.g.,\ $\tA, \tB, \tC$) for tensors of order three or greater. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor, where $K$ is the number of modes and also called the order. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as 
\[
\tY \times_1\mX_1\times \cdots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\cdots x_{j_K,i_K}^{(K)}},
\]
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use the shorthand $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as 
\[
\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}.
\]
The tensor Frobenius norm and maximum norm are defined as
\[
\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2},\quad \text{and}\quad \mnormSize{}{\tY}=\max_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}.
\]
When $a$ is a vector, we use $\vnormSize{}{a}=\langle a,a \rangle^{1/2}$ to denote the vector 2-norm.

A higher-order tensor can be reshaped into a lower-order object. We use $\textup{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ to denote the operation that reshapes the tensor along mode $k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The multilinear rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\textup{Unfold}_k(\tY)$, $k=1,\ldots,K$. We use $\sigma_{\min}(\cdot)$ and $\sigma_{\max}(\cdot)$ denote the smallest and largest matrix singular values. We let $\mI_d$ denote the $d \times d$ identity matrix and $[d]$ denote the $d$-set $\{1,\ldots,d\}$. We use $\mathbb{O}_{d,r}$ to denote the collection of all $d$-by-$r$ matrices with orthogonal columns; i.e., $\mathbb{O}_{d,r}=\{\mP\in\mathbb{R}^{d\times r}\colon \mP^T\mP = \mathbf{1}_r\}$. For ease of notation, we allow the basic arithmetic operators (e.g.,\ $+, -, \geq $) and univariate functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner. 
%We use $f'(\cdot)$, $f''(\cdot)$ to denote the first-order and second-order derivative of a function $f(\cdot)$.

@@

\subsection{Three examples}
We give three seemingly different examples that can all be formulated as our supervised tensor decomposition model~\eqref{eq:decomp}.
\begin{example}[Spatio-temporal growth model]
The growth curve model~\citep{gabriel1998generalised,srivastava2008models} was originally proposed as an example of bilinear model for matrix data, and we adopt its higher-order extension here. Let $\tY=\entry{y_{ijk}}\in\mathbb{R}^{d \times m\times n}$ denote the pH measurements of $d$ lakes at $m$ levels of depth and for $n$ time points. Suppose the sampled lakes belong to $q$ types, with $p$ lakes in each type. Let $\{\ell_j\}_{j\in[m]}$ denote the sampled depth levels and $\{t_k\}_{k\in[n]}$ the time points. Assume that the expected pH trend in depth is a polynomial of order at most $r$ and that the expected trend in time is a polynomial of order $s$. Then, the conditional mean model for the spatio-temporal growth can be represented as
\begin{equation}\label{eq:time}
\mathbb{E}(\tY|\mX_1,\mX_2,\mX_3)=\tC\times\{\mX_1\mM_1,\ \mX_2\mM_2,\ \mX_3\mM_3\},
\end{equation}
where $\mX_1=\text{blockdiag}\{\mathbf{1}_p,\ldots,\mathbf{1}_p\}\in \{0,1\}^{d\times q}$ is the design matrix for lake types, and
\[
\mX_2=
\begin{pmatrix}
1 & \ell_1&\cdots &\ell^{r}_1\\
1 & \ell_2&\cdots &\ell^{r}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&\ell_{m}&\cdots&\ell^{r}_{m}
\end{pmatrix},\quad
\mX_3=
\begin{pmatrix}
1 & t_1&\cdots &t^{s}_1\\
1 & t_2&\cdots &t^{s}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&t_{n}&\cdots&t^{s}_{n}
\end{pmatrix}
\]
are the design matrices for spatial and temporal effects, respectively, $\tC\in\mathbb{R}^{r_1\times r_2\times r_3}$ is the unknown core tensor, and $\mM_k$ are unknown dimension reduction matrices on each mode. The factors $\mX_k\mM_k$ are sufficient features in the mean model~\eqref{eq:time}. The spatial-temporal model is a special case of our supervised tensor decomposition model~\eqref{eq:decomp}, with features available on each of the three modes.
\end{example}

@@

\subsection{Connection to sufficient dimension reduction and tensor-on-tensor regression}\label{sec:connection}
One important implication of our method is that we allow high-dimensionality in both tensor dimension $d_k$ and feature dimension $p_k$. Recall that the model rank $r_k$ is typically smaller than $d_k$ and $p_k$. In such a case, the matrices $\mM_k$ serve the role of simultaneous dimension reduction of the data tensor and features. The role of the sufficient features $\mX_k\mM_k$ can be seen in the following two conditional independence assumptions in model~\eqref{eq:decomp},
\begin{align}
\tY \ci \{\mX_k\}&\ \big|\ \{\mX_k\mM_k\} \ \ \text{(independence between the tensor and features)},\\
y_{i_1,\ldots,i_K} \ci y_{i'_1,\ldots,i'_K} & \ \big|\ \{\mX_k\mM_k\}  \ \ \text{(independence within the tensor)},
\end{align}
where the second line holds for all $(i_1,\ldots,i_K)\neq(i'_1,\ldots,i'_K) \in[d_1]\times\cdots\times[d_K]$, and $\ci$ denotes the independence. The first property highlights the ``decorrelation'' role of $\mM_k$, in the same spirit as the sufficient dimension reduction~\citep{adragni2009sufficient} in supervised learning, whereas the second property highlights the tensor dimension reduction in the usual unsupervised sense (consider, for example, $\mX_k=\mathbf{I}_{d_k}$ for all $k\in[K]$).

Our model also has a close connection with tensor-on-tensor regression~\citep{raskutti2019convex, lock2018tensor,gahrooei2020multiple,hao2020sparse}. Specifically, model~\eqref{eq:decomp} can be viewed as a multivariate regression model, where the response is the vectorized tensor and the covariates are interactions between sufficient features across modes. We take an order-3 tensor under the Gaussian model for illustration. Let $\mX,\mZ,\mW$ denote the feature matrix on mode $k=1, 2, 3$, respectively. Suppose that each mode has two-dimensional sufficient features, denoted $\mM_1\mX=[\mx_1,\mx_2]$, $\mM_2\mZ=[\mz_1,\mz_2]$, $\mM_3\mW=[\mw_1,\mw_2]$. Here $\mx_1,\mx_2,\ldots,\mw_1, \mw_2$ are column vectors. Then the model~\eqref{eq:decomp} is a regression model with across-mode interactions,
\begin{equation}\label{eq:interaction}
\mathbb{E}(y_{ijk}|\mX,\mZ,\mW)=c_{111}\mx_{1i}\mz_{1j}\mw_{1k}+c_{121}\mx_{i1}\mz_{j2}\mw_{k1}+\cdots+c_{221}\mx_{2i}\mz_{2j}\mw_{1k}+c_{222}\mx_{2i}\mz_{2j}\mw_{2k},
\end{equation}
where $\entry{c_{ijk}} \in \mathbb{R}^{2\times 2\times 3}$ are unknown interaction effects, $\mx_{1i}$ denotes the $i$-th entry in the feature vector $\mx_1$, and similar notations apply to other features. Note that lower-order interactions are naturally incorporated in~\eqref{eq:interaction} if we include an intercept column in the sufficient feature matrices. Model~\eqref{eq:interaction} shows the benefit of our supervised tensor decomposition for identifying across-mode interactions.

@@

\begin{table}[htb]
\centering
\begin{tabular}{c|ccc}
Data type &Gaussian & Poisson& Bernoulli\\
\hline
Domain $\mathbb{Y}$& $\mathbb{R}$&$\mathbb{N}$&$\{0,1\}$\\
 $b(\theta)$&$\theta^2/2$& $\exp(\theta)$&$\log (1+\exp(\theta))$\\
 link $f(\theta)$&$\theta$&$\exp(\theta)$&$(1+\exp(-\theta))^{-1}$
\end{tabular}
\caption{Canonical links for common distributions.}\label{table:link}
\end{table}

In our context, we model the tensor entries $y_{i_1,\ldots,i_K}$, conditional on $\theta_{i_1,\ldots,i_K}$, as independent draws from an exponential family. Ignoring constants that do not depend on $\Theta$, the quasi log-likelihood of~\eqref{eq:decomp} is equal to Bregman distance between $\tY$ and $b'(\Theta)$:
\begin{align}\label{eq:loglikelihood}
\tL_{\tY}(\tC,\mM_1,\ldots,\mM_K)&=\langle \tY, \Theta \rangle - \sum_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K}),\\
 \text{where}\quad \Theta&=\tC\times\{\mM_1\mX_1,\ldots,\mM_K\mX_K\}.
\end{align}
We propose a constrained maximum quasi-likelihood estimator (M-estimator),
\begin{align} \label{eq:MLE} 
(\hat \tC, \hat \mM_1,\ldots,\hat \mM_k) &=\argmax_{(\tC,\mM_1,\ldots,\mM_K)\in \tP} \ \tL_{\tY}(\tC,\mM_1,\ldots,\mM_K),
\end{align}
where the parameter space $\tP$ is
\[
\tP=\left\{(\tC, \mM_1,\ldots,\mM_K) \ \Big| \ \mM_k^T\mM_k=\mathbf{I}_{r_k}\ \text{for all }k\in[K],\  \mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha \right\}.
\]
The maximum norm constraint on the linear predictor $\Theta$ is a technical condition to avoid the divergence in the non-Gaussian variance.

@@

We conclude this section by providing the accuracy, measured in KL divergence, for the response distribution.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=16cm]{compare_GLM.pdf}
\caption{Performance comparison under stochastic block models. The three panels plot the MSE when the response tensors are generated from (a) Gaussian (b) Poisson and (c) Bernoulli models. The $x$-axis represents the number of blocks in the networks. }\label{fig:glm}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=15cm]{compare_alternative.pdf} 
\caption{Comparison between different tensor methods. Panels (a) and (b) plot MSPE versus the number of modes with available features. Panels (c) and (d) plot MSPE versus the effective sample size $d^2$.
We consider rank $\mr=(3,3,3)$ (low), $\mr=(4,5,6)$ (high), and signal $\alpha =3 $ (low), $\alpha=6$ (high).}~\label{fig:compare}
\end{figure}
